---
title: "Class 12 Lab Session"
format: pdf
author: Malibu Slattery (A1844012)
toc: TRUE
---

## Proportion of G/G in a population

```{r}
mxl<-read.csv("rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```
```{r}
gbr<-read.csv("gbr_rs8067378.csv")
head(gbr)
```
Find proportion of G|G in GBR population

```{r}
table(gbr$Genotype..forward.strand.)
```
```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100,2)
```

## The variant that is associated with childhood asthma is more frequent in GBR population than in the MXL population....let's explore this further...


```{r}
expr<-read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
table(expr$geno)
```

```{r}
library(ggplot2)
```
> Q 13
  >sample size: A/A (108, median = 31.25), A/G (233, 25.06) G/G (121, 20.07). 
  
```{r}
#I didn't really know how to use the median function that was being talked about, so I used the filter function we've been working with in datacamp.
library(dplyr)

AA<-expr %>% filter(geno=="A/A")
head(AA)
round(median(AA$exp),2)

AG<-expr %>% filter(geno=="A/G")
head(AG)
round(median(AG$exp),2)

GG<-expr %>% filter(geno=="G/G")
head(GG)
round(median(GG$exp),2)


```
  
> Q.14 Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
ggplot(expr, aes(x=geno, y=  exp, fill = geno)) + geom_boxplot(notch=TRUE) 
```
> ans: A/A is more associated with the expression of ORMDL3 and G/G is less associated with the expression of ORMDL3. Thus, you are more likely to express ORMDL3 if you have the former, rather than the latter.


