---
title: "bimm143_lab5datavisR"
author: "Malibu Slattery (A18488012)"
format: html
---

## Background

There are lots of ways to make figures in R. These include so-called "base-R" graphics. (e.g. `plot()`)
and tones of add-on packages like **ggplot2**.

For example, here we make the same plot with both:
```{r}
head(cars)
plot(cars)
```
```{r}
library(ggplot2)
p<- ggplot(cars) + aes(speed, dist) + geom_smooth(col="green", method = "lm", se = FALSE) + xlab("Speed") + geom_point() + ylab("Distance") + labs(title ="Speed vs Distance Plot", subtitle = "by Malibu Slattery") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + theme(plot.subtitle =  element_text(hjust = 0.5))+ theme(plot.title = element_text(face="bold"))
p
```

## Gene expression plot
```{r}
library(ggplot2)
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
nrow(genes)
ncol(genes)
table(genes$State)
round( table(genes$State)/nrow(genes) * 100, 2 )
```

```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point() + xlab("Condition 1") + ylab("Condition 2")

#genes$State
```
Condition 2: Let's color by `State`, so we can see the "up" and "down" significant genes compared to all the "unchanging" genes..
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col = State) +
  geom_point() + xlab("Condition 1") + ylab("Condition 2")
```

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col = State) +
  geom_point() + xlab("Condition 1") + ylab("Condition 2") + scale_color_manual(values = c("darkblue", "grey", "darkgreen")) + labs(title="Gene Expression Changes upon GLP-1 Drug")+ theme(plot.title = element_text(hjust = 0.5))
```
Let's look at the famous **gapminder**!
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder, 3)

ggplot(gapminder) + aes(gdpPercap, lifeExp, col = continent) + geom_point(alpha = 0.3) + xlab("GDP per Capita") + ylab("Life Expectancy") + guides(color = guide_legend(title = "Continent"))
```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess above!
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder, 3)

ggplot(gapminder) + aes(gdpPercap, lifeExp, col = continent) + geom_point(alpha = 0.3) + xlab("GDP per Capita") + ylab("Life Expectancy") + facet_wrap(~continent) + guides(color = guide_legend(title = "Continent"))
```






