---
title: "class18"
format: pdf
toc: true
author: Malibu Slattery (A18488012)
---

## Pertussis and the CMI-PB project

# Background
Pertussis/whooping cough is a lung infection caused by bacteria B. It can infect anyone but is most deadly for infants under 1 year of age.

#CDC tracking data
The CDC

```{r, echo =FALSE}
cdc <- 
  data.frame(
                            year = c(1922L,1923L,1924L,1925L,1926L,1927L,
                                     1928L,1929L,1930L,1931L,1932L,1933L,1934L,
                                     1935L,1936L,1937L,1938L,1939L,1940L,
                                     1941L,1942L,1943L,1944L,1945L,1946L,
                                     1947L,1948L,1949L,1950L,1951L,1952L,
                                     1953L,1954L,1955L,1956L,1957L,1958L,
                                     1959L,1960L,1961L,1962L,1963L,1964L,
                                     1965L,1966L,1967L,1968L,1969L,1970L,
                                     1971L,1972L,1973L,1974L,1975L,1976L,1977L,
                                     1978L,1979L,1980L,1981L,1982L,1983L,
                                     1984L,1985L,1986L,1987L,1988L,1989L,
                                     1990L,1991L,1992L,1993L,1994L,1995L,
                                     1996L,1997L,1998L,1999L,2000L,2001L,
                                     2002L,2003L,2004L,2005L,2006L,2007L,
                                     2008L,2009L,2010L,2011L,2012L,2013L,
                                     2014L,2015L,2016L,2017L,2018L,2019L,2020L,
                                     2021L,2022L,2023L, 2024L, 2025L),
    cases = c(107473,164191,165418,152003,202210,181411,
                                     161799,197371,166914,172559,215343,
                                     179135,265269,180518,147237,214652,
                                     227319,103188,183866,222202,191383,191890,
                                     109873,133792,109860,156517,74715,
                                     69479,120718,68687,45030,37129,60886,
                                     62786,31732,28295,32148,40005,14809,
                                     11468,17749,17135,13005,6799,7717,9718,
                                     4810,3285,4249,3036,3287,1759,2402,
                                     1738,1010,2177,2063,1623,1730,1248,
                                     1895,2463,2276,3589,4195,2823,3450,4157,
                                     4570,2719,4083,6586,4617,5137,7796,
                                     6564,7405,7298,7867,7580,9771,11647,
                                     25827,25616,15632,10454,13278,16858,
                                     27550,18719,48277,28639,32971,20762,
                                     17972,18975,15609,18617,6124,2116,3044,
                                     7063, 22538, 21996)
  )
```


> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time. 

```{r}
library(ggplot2)

ggplot(cdc, aes(x=year, y = cases)) + geom_point() + geom_line() + labs(title = "Year vs Number of Whooping Cough Cases")

```

# A tale of two vaccines (wP & aP)

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc, aes(x=year, y = cases)) + geom_point() + geom_line() + labs(title = "Year vs Number of Whooping Cough Cases") + geom_vline(xintercept=1946, col = "blue")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

```{r}
ggplot(cdc, aes(x=year, y = cases)) + geom_point() + geom_line() + labs(title = "Year vs Number of Whooping Cough Cases") + geom_vline(xintercept=1946, col = "blue", lty = 2) + geom_vline(xintercept=1996, col = "red", lty = 2) + geom_vline(xintercept=2020, col = "grey", lty = 2) 
```

The initial decrease after 1946 was from the inital strong reaction and protection against the virus. On the other hand, after 1996, less people may be getting vaccinated. Waning efficacy of the aP vaccine. There seems to be a 10 year lag in spikes of cases.

## Exploring CMI-PB data

# The CMI-PB API returns JSON data

The mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

They made their data available in the format API. We can read this into R with the `read_json()` function from the **jsonlite** package.

```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```

```{r}
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```


> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)? eg:`biological_sex` and `race`...is this representitve of the US populatio

```{r}
table(subject$race, subject$biological_sex)
```

# Side-Note: Working with dates 

```{r}
library(lubridate)
today()
```

```{r}
today() - ymd("2000-01-01")
```


```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(dplyr)

subject$age <- today() - ymd(subject$year_of_birth)

wp <- subject |> filter(infancy_vac=="wP")

round( summary( time_length( wp$age, "years" ) ) )
```
```{r}
ap <- subject |> filter(infancy_vac=="aP")

round( summary( time_length( ap$age, "years" ) ) )
```

They are significantly different because there may be more available data to test the vaccines on younger individuals.

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different? 

```{r}
ggplot(subject) +
  aes(time_length(age, unit="year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=TRUE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")

x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

# Joining multiple tables

We can read more tables from the CBI-PB database

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector=T)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector =T)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```
make sense of all this data we need to "join" (aka merge or link) all these tables together. Only then will you know that a given Ab measurement(from the `ab_titer` table) was collected on a certain date from the `specimen` table from a certain cP or aP subject from the `subject` table.

We can use **dplyr** and the `*_join()` family of functions to do this.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```
Let's do one more `inner_join` to join the ab_titer with `meta` data...

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata<-inner_join(ab_titer, meta)
head(abdata)
```
> Q11. How many different Ab isotype values are there in this dataset?

```{r}
sometable<-table(abdata$isotype, abdata$infancy_vac)
```


> Q12. How many different "antigen" values are measured? 

```{r}
table(abdata$antigen)
```
Let's focus on IgG isotype...

## Examine IgG Ab titer levels

```{r}
igg <-abdata |> 
          filter(isotype=="IgG")
#igg
```

Make a plot of `MFI_normalised` values for all `antigen` values.

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg, aes(x=MFI_normalised, y = antigen)) + geom_boxplot()
```
The antigens `PT` and `FIM2/3` AND `FHA` appear to have the widest range of values. 

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

```{r}
ggplot(igg, aes(x=MFI_normalised, y = antigen)) + geom_boxplot() + facet_wrap(~ infancy_vac)
```

Is there a difference with time (i.e. before booster vs after booster)?

```{r}
ggplot(igg, aes(x=MFI_normalised, y = antigen, col = infancy_vac)) + geom_boxplot() + facet_wrap(~ visit)
```
```{r}
igg |> filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```
> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") |>
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
```{r}
filter(igg, antigen=="FIM2/3") |>
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
> Q16. What do you notice about these two antigens time courses and the PT data in particular?
      ans: There is a peak in PT data at visit 4-5, but goes down significantly after that, showing how much more effective it is than the OVA.


> Q17. Do you see any clear difference in aP vs. wP responses?
      ans: Not extreme differences, but the wP responses seem to be higher in days 5 and 6.


```{r, message = FALSE}
ab.PT.21 <- abdata |>
    filter(dataset=="2021_dataset", isotype == "IgG", antigen == "PT")

  ggplot(ab.PT.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
        geom_point()+
    geom_smooth(aes(x = planned_day_relative_to_boost, y = MFI_normalised, group = infancy_vac), se= FALSE, , size = 1.5, span = .3) +
    geom_line(alpha = .3) +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="CMI-PB 2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") + xlab("Day Relative to Boost") + ylab("Normalised MFI")
```
https://r02pro.github.io/smoothline.html 
https://bookdown.dongzhuoer.com/hadley/ggplot2-book/plot-geoms 

> Q18. Does this trend look similar for the 2020 dataset?
    ans: yes, it does.

# Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

> Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```
> Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?
      ans: the max level is at the 4th visit

> Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?
      ans: this does match the trend of the antibody titer data because the response/expression is highest at the 4th visit.
      
```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```
```{r}
ssrna %>%  
  filter(visit==4) |>
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```
      
